---
title: "_GDSArray_: GDS back-end of DelayedArray"
author:
- name: Qian Liu
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Hervé Pagès
  affiliation: Fred Hutchinson Cancer Research Center, Seattle, WA
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
date: "last edit: 3/1/2018"
output:
  BiocStyle::html_document:
      toc: true
      toc_float: true
package: GDSArray
vignette: |
  %\VignetteIndexEntry{GDSArray: GDS back-end of DelayedArray}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

GDSArray is a Bioconductor package that represents GDS back-end of DelayedArray. It could convert the GDS nodes in a DelayedArray data structure. The rich common methods and data operations defined on `GDSArray` makes it more R-user-friendly. The array data from GDS file are always returned with the first dimension being `variants/snps` and the second dimension being `samples`. This feature is consistent with the array data saved in `SummarizedExperiment`, and makes the `GDSArray` package more interoperable with other established Bioconductor data infrastructure. 

# Package installation

1. Download the package from Bioconductor. 

```{r getPackage, eval=FALSE}
source("https://bioconductor.org/biocLite.R")
biocLite("GDSArray")
```

2. Load the package into R session.
```{r Load}
library(GDSArray)
```

```{r options, eval=FALSE, echo=FALSE}
options(showHeadLines=3)
options(showTailLines=3)
```

# GDS format introduction

## Genomic Data Structure (GDS)
The Bioconductor package `gdsfmt` has provided a high-level R interface to CoreArray Genomic Data Structure (GDS) data files, which is designed for large-scale datasets, especially for data which are much larger than the available random-access memory.  
The GDS format has been widely used in genetic/genomic research for high-throughput genotyping or sequencing data. There are two major classes that extends the `gds.class`: `SNPGDSFileClass` suited for genotyping data (e.g., GWAS), and `SeqVarGDSClass` that are designed specifically for DNA-sequencing data. The file format attribute in each data classes are set as `SNP_ARRAY` and `SEQ_ARRAY`. There are rich functions written based on these data classes for common data operation and statistical analysis.  
More details about GDS format could be found here: [gdsfmt](https://bioconductor.org/packages/release/bioc/vignettes/gdsfmt/inst/doc/gdsfmt_vignette.html), [SNPRelate](http://bioconductor.org/packages/release/bioc/vignettes/SNPRelate/inst/doc/SNPRelateTutorial.html), and [SeqArray](http://bioconductor.org/packages/release/bioc/vignettes/SeqArray/inst/doc/SeqArrayTutorial.html)


# GDSArray, GDSMatrix 

GDSArray is the GDS back-end of DelayedArray. It has methods like `dim`, `dimnames` defined, and it inherites array-like operations and methods from DelayedArray, e.g., the subsetting method of `[`.

## Available GDS nodes 
`gdsNodes` is an exported function, which takes the GDS file path as input, and return all available nodes that are convertable to `GDSArray`. The returned GDS node names will be used as input for the `GDSArray` constructor. 

```{r, gdsNodes}
file <- SeqArray::seqExampleFileName("gds")
gdsNodes(file)
```

## GDSArraySeed

The `GDSArraySeed` is the seed for the `GDSArray` object, which is not exported from this package. Seed objects should contain the GDS file path, and are expected to satisfy the “seed contract” i.e. to support dim() and dimnames().

```{r, GDSArraySeed}
seed <- GDSArray:::GDSArraySeed(file, "genotype/data")
seed
```

## GDSArray, GDSMatrix, and DelayedArray

The `GDSArray` constructor over a single argument of `GDSArraySeed` will return the `GDSArray` object directly. Or it could take the arguments of filepath, and the GDS node inside the GDS file. The following 2 lines of code returns the same `GDSArray` object.  
One thing to note, the `GDSArray` constructor always return the object with rows being "genes / variants / snps" and the columns being "samples", which is consistent with the assay data inside `SummarizedExperiment`. 

```{r, GDSArray}
GDSArray(seed)
GDSArray(file, "genotype/data")
```
The `GDSMatrix` is the 2-dimensional `GDSArray`, and will be returned from the `GDSArray` constructor automatically if the input GDS node is 2-dimensional. 

```{r, GDSMatrix}
GDSArray(file, "phase/data")
```

The `DelayedArray` constructor over `GDSArraySeed` object will return the same content as `GDSArray` constructor over the same `GDSArraySeed`. 

```{r, da}
class(DelayedArray(seed))
``` 

## `GDSArray` methods

### dim, dimnames
The `dimnames(GDSArray)` returns an unnamed list, with the length of each element to be the same as return from `dim(GDSArray)`. 
```{r, dims}
ga <- GDSArray(file, "phase/data")
dim(ga)
class(dimnames(ga))
lengths(dimnames(ga))
```

### `[` subsetting

```{r, methods}
ga[1:3, 10:15]
ga[c(rep(c(TRUE, FALSE), 4), rep(FALSE, nrow(ga)-8)), ]
```

# sessionInfo
```{r, sessionInfo}
sessionInfo()
```

